---
title: "R Day 2"
author: "Yucheng Zhu"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
here::here("data","gapminder5.csv")

gapminder <- read.csv(here::here("data","gapminder5.csv"))
head(gapminder)
str(gapminder)
```
## Converting factors to characters
```{r}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)

mean(gapminder$lifeExp[gapminder$country == "Afghanistan"])
mean(gapminder$lifeExp[gapminder$country == "Albania"])

```
```{r}
obs <- 1:nrow(gapminder)
for (i in obs)
  {
  gapminder[i,"gdp"] <- gapminder[i,"pop"] * gapminder[i,"gdpPercap"]
}
```

## Creating logged variables with a for loop
```{r}
for (i in obs)
{
  gapminder[i,"log_gdpPercap"] <- log(gapminder[i,"gdpPercap"])
  gapminder[i,"log_pop"] <- log(gapminder[i,"pop"])
}
head(gapminder)


```
```{r}
years <- unique(gapminder$year)

for (i in years)
{
  mean_le <- mean(gapminder$lifeExp[gapminder$year==i],
                  na.rm = T)
  print(paste0(i,": ", mean_le))
}

```

```{r}
conts <- unique(gapminder$continent)
for (i in conts)
{
  mean_le <- mean(gapminder$lifeExp[gapminder$continent == i],
                  na.rm = T)
  print(paste0(i, ":", mean_le))
}
```

```{r}
for (i in conts)
{
  print(paste0("Continent:", i))
  for (j in years)
  {
    sd_le <- sd(gapminder$lifeExp[gapminder$continent == i &
                                    gapminder$year == j],
                na.rm = T)
      print(paste0(j, ":", sd_le))
  }
}
```


```{r}
i <- 1987

while (i <= 2002)
{
  sd_le <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i,": ", sd_le))
  i <- i+5
}

```



```{r}
threshold <- 1987

for (i in unique(gapminder$year))
{
  tmp <- mean(gapminder$lifeExp[gapminder$year == i])
  
  if ( i >= threshold)
  {
    print(paste("Mean Life Exp in", i, "is", tmp))
  }
  else
  {
    print(paste("Condition not meet"))
  }
}

```


```{r}

report_stats <-
  function(df, variable, continent)
  {
    var <- gapminder[[variable]][gapminder$continent == continent]
    min_le <- min(var)
    max_le <- max(var)
    mean_le <- mean(var)
    median_le <- median(var)
    cat("continent:", continent,
        "\nMin LE:", min_le,
        "\nMax LE:", max_le,
        "\nMean LE:", mean_le,
        "\nMedian LE:", median_le)
  }
report_stats(gapminder, "lifeExp", "Asia")


```


## Excercises Day 2

---
title: "Exercises Day 2"
author: "Richard Paquin Morel, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: FALSE
---


```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Load the `gapminder` dataset.

```{asis}
gapminder <- read.csv((here::here("data/gapminder5,csv", stringsAsFactors=FALSE)))
```

```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```


## If Statement

Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the `any` function.

```{asis}
### Answer
year<- 2002
if(any(gapminder$year == year))
{
  print(paste("Record for", year,"found"))
}
else
{
  print(paste("No recordes for year",year))
}
```

```{r}
year<-2002
if(any(gapminder$year == year)){
   print(paste("Record(s) for the year",year,"found."))
} else {
  print(paste("No records for year",year))
}
```


## Loop and If Statements

Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

Write a script that loops through the `gapminder` data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{asis}
### Answer

overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country))
{
  country_mean <- mean(gapminder$pop[gapminder$country == i])
  if (country_mean < overall_mean){
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i])
    print(paste("Mean Life Exp in", i, "is", mean_le))
      }
}


```

```{r}
overall_mean <- mean(gapminder$pop)

for (i in unique(gapminder$country)) {
   country_mean <- mean(gapminder$pop[gapminder$country==i])
   
   if (country_mean < overall_mean) {
     mean_le <- mean(gapminder$lifeExp[gapminder$country==i])
     print(paste("Mean Life Expectancy in", i, "is", mean_le))
   } 
} # end for loop
```

```{r}
lower_threshold <- 50
upper_threshold <- 70
 
for (i in unique(gapminder$continent)){
    tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
    
    if (tmp < lower_threshold){
        print(paste("Average Life Expectancy in", i, "is less than", lower_threshold))
    }
    else if (tmp > lower_threshold & tmp < upper_threshold){
        print(paste("Average Life Expectancy in", i, "is between", lower_threshold, "and", upper_threshold))
    }
    else {
        print(paste("Average Life Expectancy in", i, "is greater than", upper_threshold))
    }
   
}
```


## Exercise: Write Functions

Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset.  Hint: Use `mode()` or `class()` to get the class of the data in each column. Remember that `names()` or `colnames()` returns the name of the columns in a dataset.

```{asis}
### Answer
data_frame_info <- function(df){
  cols <- names(df)
  for (i in cols) {
    print(paste(i,":",mode(df[,i])))
    
  }
}
Note: Some of these were taken or modified from https://www.r-bloggers.com/functions-exercises/
```

```{r}
data_frame_info <- function(df) {
  cols <- names(df)
  for (i in cols) {
    print(paste0(i, ": ", mode(df[, i])))
  }
}
data_frame_info(gapminder)
```

Create a function that given a vector will print the mean and the standard deviation of a **vector**, it will optionally also print the median. Hint: include an argument that takes a boolean (`TRUE`/`FALSE`) operator and then include an `if` statement.

```{asis}
### Answer
vector_info <- function(x, include_median = FALSE){
  print(paste(mean(x)))
  print(paste(sd(x)))
  if (include_median){
    print(paste(median(x)))
  }
}









```

```{r}
vector_info <- function(x, include_median=FALSE) {
  print(paste("Mean:", mean(x)))
  print(paste("Standard Deviation:", sd(x)))
  if (include_median) {
    print(paste("Median:", median(x)))
  }
}

le <- gapminder$lifeExp
vector_info(le, include_median = F)
vector_info(le, include_median = T)
```

## Analyzing the relationship 

Use what you've learned so far to answer the following questions using the `gapminder` dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)




2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.









